---
output: github_document
---

<!-- README.md is generated from README.Rmd. -->

# RClinVarbitration

[![R-CMD-check](https://github.com/RGenomicsETL/RClinVarbitration/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/RGenomicsETL/RClinVarbitration/actions/workflows/R-CMD-check.yaml)
[![pkgdown](https://github.com/RGenomicsETL/RClinVarbitration/actions/workflows/pkgdown.yaml/badge.svg)](https://rgenomicsetl.github.io/RClinVarbitration/)
[![r-universe](https://rgenomicsetl.r-universe.dev/badges/:name)](https://rgenomicsetl.r-universe.dev/)

`RClinVarbitration` joins ClinVar's official `variant_summary` and
`submission_summary` reports directly into one tidy DuckDB table. Each
decision, SCV submission, RCV accession, and allele-gene link is one scalar
row. The package applies the ClinVarbitration decision policy during the same
import; no XML database is required for ordinary release tracking.

VCV XML import remains available when an analysis needs source observations,
citations, or text that the flat reports do not carry.

The package bundles exact-version DuckDB extensions for DuckDB `v1.5.0` through
`v1.5.5`. Connections must allow locally built unsigned extensions.

## Installation and platforms

```{r installation, eval=FALSE}
install.packages(
  "RClinVarbitration",
  repos = c(
    rgenomicsetl = "https://rgenomicsetl.r-universe.dev",
    CRAN = "https://cloud.r-project.org"
  )
)
```

Native builds support Linux, macOS, and x86-64 Windows. Linux and macOS require
libxml2, zlib, `pkg-config`, and a C compiler. On Windows, install a current
Rtools release; its target-aware `pkg-config` supplies the static libxml2 and
zlib dependencies. webR is supported separately through the tested Emscripten
build.

## Quick start

```{r quick-start}
library(DBI)
library(duckdb)
library(RClinVarbitration)

con <- dbConnect(duckdb())
variant_report <- system.file(
  "extdata", "variant_summary_fixture.txt",
  package = "RClinVarbitration"
)
submission_report <- system.file(
  "extdata", "submission_summary_fixture.txt",
  package = "RClinVarbitration"
)

imported <- rclinvarbitration_import_flat(
  con,
  submission_report,
  variant_report,
  release_id = "clinvar-example"
)

dbGetQuery(con, "
  SELECT record_kind, count(*) AS rows
  FROM clinvar
  GROUP BY record_kind
  ORDER BY record_kind
")
```

For a complete current or archived release, download both reports and use a
file-backed database:

```{r full-release, eval=FALSE}
release <- "2026-03"
reports <- rclinvarbitration_download_clinvar(
  release,
  file = c("submission_summary", "variant_summary")
)

full_con <- dbConnect(duckdb(
  dbdir = "clinvar.duckdb",
  config = list(
    memory_limit = "8GB",
    preserve_insertion_order = "false",
    threads = "4"
  )
))
full_import <- rclinvarbitration_import_flat(
  full_con,
  reports[["submission_summary"]],
  reports[["variant_summary"]],
  release_id = paste0("ncbi-clinvar-", release),
  parquet_path = "clinvar-2026-03.parquet"
)
```

The table is deliberately long rather than nested:

| `record_kind` | One row represents |
|:--|:--|
| `variation`, `allele` | a ClinVar variation or constituent allele |
| `location` | one assembly-specific VCF placement |
| `decision` | the allele-level ClinVarbitration decision |
| `scv_assertion` | one submitted classification |
| `rcv_assertion` | one RCV accession and phenotype |
| `gene` | one allele-gene link |

GRCh37 and GRCh38 locations are retained together. `clinvar_vcf` exposes
one-based `position`, `reference`, and `alternate` fields with conventional
`1`/`chr1` primary-contig names. The exact `sequence_accession` is retained;
alternate placements use that accession as `contig`, and X/Y PAR placements
remain separate rows. Raw `location` rows are still retained when ClinVar lacks
a complete VCF tuple, as occurs for some structural variants; those rows are
deliberately absent from `clinvar_vcf`.

Measured release summaries are rendered from their committed receipts:

```{r benchmark-summary, echo=FALSE}
receipt_path <- function(name) {
  candidates <- c(
    file.path("inst", "benchmarks", name),
    file.path("..", "inst", "benchmarks", name),
    system.file("benchmarks", name, package = "RClinVarbitration")
  )
  candidates[file.exists(candidates)][1L]
}
read_receipt <- function(name) as.list(read.dcf(receipt_path(name))[1L, ])
format_count <- function(value) {
  format(as.numeric(value), big.mark = ",", scientific = FALSE, trim = TRUE)
}
format_gib <- function(value) sprintf("%.2f GiB", as.numeric(value) / 1024^3)

flat_receipt <- read_receipt("full-flat-release-2026-03.dcf")
xml_receipt <- read_receipt("full-release-2026-07-02.dcf")
benchmark_summary <- data.frame(
  Source = c(
    "[March 2026 flat](inst/benchmarks/full-flat-release-2026-03.dcf)",
    "[2 July 2026 XML](inst/benchmarks/full-release-2026-07-02.dcf)"
  ),
  `Stored facts` = c(
    format_count(flat_receipt$total_rows),
    format_count(xml_receipt$total_rows)
  ),
  DuckDB = c(
    format_gib(flat_receipt$database_bytes),
    format_gib(xml_receipt$database_bytes)
  ),
  `GRCh38 raw / VCF` = c(
    paste(
      format_count(flat_receipt$raw_location_rows_grch38),
      format_count(flat_receipt$vcf_rows_grch38),
      sep = " / "
    ),
    paste(
      format_count(xml_receipt$raw_location_rows_grch38),
      format_count(xml_receipt$vcf_rows_grch38),
      sep = " / "
    )
  ),
  `GRCh37 raw / VCF` = c(
    paste(
      format_count(flat_receipt$raw_location_rows_grch37),
      format_count(flat_receipt$vcf_rows_grch37),
      sep = " / "
    ),
    paste(
      format_count(xml_receipt$raw_location_rows_grch37),
      format_count(xml_receipt$vcf_rows_grch37),
      sep = " / "
    )
  ),
  check.names = FALSE
)
knitr::kable(benchmark_summary, align = c("l", "r", "r", "r", "r"))
```

XML adds observations, citations, names, cross-references, attributes, and
text. Both paths preserve assembly and sequence-accession identity.

```{r coordinate-view}
dbGetQuery(con, "
  SELECT assembly, contig, position, reference, alternate
  FROM clinvar_vcf
  ORDER BY assembly
")
```

## DuckLake publication and release changes

The [RGenomicsETL `ducklake-r`
fork](https://github.com/RGenomicsETL/ducklake-r) registers that Parquet
without collecting it into R. RClinVarbitration owns the key-based publication:

```{r ducklake-publication, eval=FALSE}
ducklake::set_ducklake_connection(full_con)
ducklake::attach_ducklake("clinvar_lake", lake_path = "clinvar-lake")
publication <- rclinvarbitration_publish_ducklake(full_con, full_import)
changes <- ducklake::get_table_changes(
  "clinvar",
  publication$snapshot_id,
  publication$snapshot_id
)
```

The function initializes persistent staging once, validates keys and policy
identity, and commits inserts, updates, and withdrawals as one snapshot.
Unchanged rows are untouched. DuckLake's change feed is the delta authority.

Complete-release row counts and storage depend on the selected source path:
the compact flat reports contain the ordinary arbitration substrate, while XML
adds source entities that are absent from those reports. The storage vignette
records the measured workloads separately rather than treating them as the
same benchmark.

Read the [arbitration algorithm](https://rgenomicsetl.github.io/RClinVarbitration/articles/arbitration-algorithm.html),
[storage and caching guide](https://rgenomicsetl.github.io/RClinVarbitration/articles/storage-cache-and-performance.html),
and [semantic/DuckLake/VariantStory integration](https://rgenomicsetl.github.io/RClinVarbitration/articles/semantic-ducklake-variantstory.html).
The [deviation and differential audit](https://github.com/RGenomicsETL/RClinVarbitration/blob/main/docs/ERRATA.md) records known differences
from upstream ClinVarbitration and ClinVar.

## Submitter exclusions

Flat imports retain all source submissions. Exclusions change only the
`decision` rows and are recorded in the import receipt. Names are matched
case-insensitively after trimming whitespace.

```{r exclusions, eval=FALSE}
rclinvarbitration_import_flat(
  con, submission_report, variant_report,
  release_id = "clinvar-example",
  submitter_exclusions = c("Example laboratory", "Another submitter")
)
```

## Optional XML enrichment

Call `rclinvarbitration_enable()`, `rclinvarbitration_init()`, and
`rclinvarbitration_import_xml()` only when the analysis needs XML-only
observations, citations, attributable text, or disease-scoped assertion
structure. The XML path writes the same scalar `clinvar` table as the flat
import. Compatibility relation names such as `clinvar_locations` and
`clinvar_scv_assertions` are views over that table. XML-derived policy
decisions remain views until an explicit Parquet or DuckLake publication asks
to materialize them.

## Comparison with upstream ClinVarbitration

The policy is pinned to [Centre for Population Genomics
ClinVarbitration](https://github.com/populationgenomics/clinvarbitration)
2.2.11 at commit `658b9f241eb2d43aa11214b153b19c1e18a16337`.

| | Upstream 2.2.11 | RClinVarbitration |
|:--|:--|:--|
| Primary input | NCBI submission and variant summary files | the same two reports |
| Runtime | Python, Hail, Nextflow, bcftools | R, DuckDB, package-owned C extension |
| Decision scope | allele | allele; optional XML disease enrichment |
| Main outputs | TSV, Hail Table, VCF, PM5 resource | one tidy DuckDB/Parquet relation |
| Submitter exclusion | `site_blacklist` / `-b` | `submitter_exclusions` or named profiles |
| PM5 | included | out of scope |

The shared decision rules include the 2016 ACMG date filter, classification
bins, 60/20 majority rule, strong-review precedence, and star calculation.
The compatibility export retains the upstream seven-column decision schema.
An exact-input execution
of the pinned upstream Python TSV stage and this package's flat reproducer over
the complete March 2026 archives produced the same 4,125,389 keys with zero
classification or star differences. Input, code, configuration, and output
digests are in the
[oracle manifest](https://github.com/RGenomicsETL/RClinVarbitration/blob/main/inst/audits/march-2026-flat-exact-oracle.dcf).

The independently matched XML/flat audit classified every one of the 16 shared
value differences and 361 key-set differences with source-row receipts. Most
come from NCBI flat rows whose classification is `-` while XML carries a current
classification; the remainder are one duplicate-SCV identity case, five source
vocabulary differences, and two nested compound alleles. The
[ERRATA audit](https://github.com/RGenomicsETL/RClinVarbitration/blob/main/docs/ERRATA.md)
contains the full counts and receipts. A published Zenodo release with 16,865
reference-only keys used an unpinned source snapshot; it is not the exact-input
conformance result.

One deliberate edge-case difference is that RClinVarbitration applies the
qualified Illumina benign exclusion declared by upstream. At the pinned commit,
the Python implementation's inner-loop `continue` does not actually remove that
submission, so compatibility here follows the documented policy rather than
that implementation accident.

## Evidence retrieval and reanalysis

A second motivation is to make ClinVar evidence reusable, not only to annotate
known alleles. The flat table retains submission descriptions, phenotype
strings, RCVs, and gene links. Optional XML enrichment adds citations,
normalized HPO links, and other attributable observations. These are joinable
inputs for
[`ducksemantics`](https://github.com/RGenomicsETL/ducksemantics), DuckLake
release history, and the source-observation model planned by
[`VariantStory`](https://github.com/RGenomicsETL/VariantStory).

These workflows retrieve and rank evidence; an embedding neighbor does not
classify a variant. Provider identity, release, source rows, deterministic
evidence admission, and human review must remain explicit.

```{r cleanup, include=FALSE}
dbDisconnect(con, shutdown = TRUE)
```

## Acknowledgements

The decision policy is adapted from Centre for Population Genomics
ClinVarbitration 2.2.11 under its MIT license. ClinVar source data are provided
by NCBI.
